Spike spectra for recurrences

### 2022-08-03 Manuscript/inter_spike_manuscript_mdpi.tex

The signal, which we would like to transform is decomposed into a set of appropriate basis functions.\footnote{We focus on $\tau$-RR in this article, Eq.~\eqref{eq_tau_rr}, since 
such time series are often spiky, which favors the proposed decomposition technique. But the proposed method can be applied to any sort of signal.} Instead of using trigonometric 
functions, as it is the idea in the Fourier decomposition, we use Dirac combs (DC) with 
different inter-spike periods as basis functions, Eq.~\eqref{eq_dirac_comb}. Let $s(t_i)$ be the normalized signal we want to transform of length $N$ and 
$t_i=i\cdot \Delta t,~i=1,\ldots,N$, where $\Delta t$ denotes the sampling time and $s(t_i) \in [0,\ 1]\ \forall\ i$. In the following we label this time series as a $(1\times N)$-dimensional 
vector $\vec{s}$. 
First, $\tilde{N}$ different DCs of length $N$ are constructed with inter-spike periods $T_\text{is} \in [1,\ldots,\tilde{N}]$ and $\tilde{N}=\ceil*{N/2}+1$. Second, in order to account 
for possible phase shifts of 
these basis functions occurring in $\vec{s}$, each of these $\tilde{N}$ different DCs also need to be shifted one step further $T_\text{is}-1$ times. This leaves us with a total number of 
$M = \sum_{i=1}^{\tilde{N}}i$ 
basis functions which we can arrange as rows of a $(M\times N)$-sized matrix $\bfX$
(Fig.~\ref{fig_tau_rr_basis_functions} illustrates the described procedure):
\begin{align}
\label{eq_basis_matrix} 
\bfX_{i,j} &= \sum_{k=0}^{N}\delta\left(j-1-kT_\text{is}(i)-i+T_\text{is}(i) \right), \quad i=1,\dots,\ceil*{N/2}+1, \quad j=1,\dots,N\\
\label{eq_T_is} T_\text{is}(i)&= n, \quad \forall n:~ \frac{n(n-1)}{2} + 1 \leq i <  \frac{n(n+1)}{2} + 1,\quad n \in \mathbb{N}_+.
\end{align}
Note that due to the shifting of each of the basis functions of inter-spike period $T_\text{is}$, $\bfX$ is not linear independent anymore. 
Furthermore, there will be identical basis functions and also basis functions, which do not allow for an unambiguous inter-spike period, 
if we would include all $N$ possible inter-spike periods for a signal of length $N$ instead of $\ceil*{N/2}+1$ (Fig.~\ref{fig_tau_rr_basis_functions}A). 
The reason is that in contrast to a trigonometric decomposition, where the Nyquist frequency marks a lower bound for the corresponding wave period, here 
the maximum considered inter-spike period is bounded by $T_{\text{is}}^{\text{max}} = \ceil*{N/2}+1$ (schematically illustrated in 
Fig.~\ref{fig_tau_rr_basis_functions}B).\\

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Inter-spike spectra of power grid frequency data}

@@

A novel type of powerspectrum, the \textit{inter-spike spectrum}, has been proposed. The method decomposes any arbitrary signal into basis functions which consist of (lagged) Dirac 
combs (DC) of different inter-spike period. The loading for each period is obtained by a regularized regression, which promotes sparsity in its solution (we chose LASSO or a sequentially 
thresholded least squares regression STLS). Since there are 
$M = \sum_{i=1}^{\ceil*{N/2}+1}i$ basis functions for a signal of length $N$ the regression can get computationally intensive for $N>1,000$. When plotting the computed loadings as a function 
of the period (or frequency) the inter-spike spectrum is obtained. This novel powerspectrum is superior to an ordinary FFT-based powerspectrum, when the signal has a spike-train-like 
appearance. Due to the sparse regression underlying the method, there is no unique inverse of the transformation and the regularization parameter plays a crucial role and determines the 
appearance of the obtained inter-spike spectrum. Moreover, similar to the Nyquist frequency barrier in the Fourier Transform which sets a lower bound for the corresponding wave period, here 
the maximum considered inter-spike period is bounded by $T_{\text{is}}^{\text{max}} = \ceil*{N/2}+1$.

The invention of the proposed method has been motivated by the idea of transforming $\tau$-recurrence rate signals ($\tau$-RRs) into their frequency domain. 
This general idea \cite{Zbilut2008} allows for a frequency analysis of high dimensional systems, because the RP is a representation of the system's state space trajectory.   
The $\tau$-RR of a recurrence plot (RP) usually has a spiky shape, especially for map-like data, and the inter-spike spectrum can reliably reveal the system's dominant frequencies, 
which is not possible when Fourier transforming the $\tau$-RR or the underlying signal itself. Since the position of the peaks 
in the $\tau$-RR are not sensitive to noise, the corresponding inter-spike spectrum also yields robust results in the presence of noise. We have successfully used the idea 
of transforming the $\tau$-RR for the detection of bifurcations in the Logistic map. By constructing appropriate surrogates of the inter-spike spectra, and thus a null model, 
the number of significant peaks in the inter-spike spectrum correlated well with the positive Lyapunov exponent. This measure was also able to resolve period-doubling bifurcations. 
\todo[inline]{Here a quick summary of the other applications.}

@@

\section{Inter-spike spectra for noisy R\"ossler system}

\begin{figure}[h!]
 \centering
 \includegraphics[width=0.9\textwidth]{./figures/fig_tau_rr_example_roessler_noise}
 \caption{Same as in Fig.~\ref{fig_tau_rr_example_roessler}, but here with 5\% additive Gaussian white noise on each component $x$, $y$ and $z$. The appearance of the inter-spike spectra 
 in \textbf{J, K, L}, and the Fourier spectra in \textbf{M, N, P} are unaffected by the additive noise.}
\label{fig_tau_rr_example_roessler_noise}
\end{figure}

@@

>>>>

The signal, which we would like to transform is decomposed into a set of appropriate basis functions.\footnote{We focus on $\tau$-RR in this article, Eq.~\eqref{eq_tau_rr}, since 
such time series are often spiky, which favors the proposed decomposition technique. But the proposed method can be applied to any sort of signal.} Instead of using trigonometric 
functions, as it is the idea in the Fourier decomposition, we use Dirac combs (DC) with 
different inter-spike periods as basis functions, Eq.~\eqref{eq_dirac_comb}. Let $s(t_i)$ be the normalized signal we want to transform of length $N$ and 
$t_i=i\cdot \Delta t,~i=1,\ldots,N$, where $\Delta t$ denotes the sampling time and $s(t_i) \in [0,\ 1]\ \forall\ i$. In the following we label this time series as a $(1\times N)$-dimensional 
vector $\vec{s}$. 
First, $\tilde{N}$ different DCs of length $N$ are constructed with inter-spike periods $T_\text{is} \in [1,\ldots,\tilde{N}]$ and $\tilde{N}=\ceil*{N/2}+1$. Second, in order to account 
for possible phase shifts of 
these basis functions occurring in $\vec{s}$, each of these $\tilde{N}$ different DCs also need to be shifted one step further $T_\text{is}-1$ times. This leaves us with a total number of 
$M = \sum_{i=1}^{\tilde{N}}i$ 
basis functions which we can arrange as rows of a $(M\times N)$-sized matrix $\bfX$
(Fig.~\ref{fig_tau_rr_basis_functions} illustrates the described procedure):
\begin{align}
\label{eq_basis_matrix} \bfX_{i,j} &= \sum_{k=0}^{N}\delta\left(j-1-kT_\text{is}(i)-i+T_\text{is}(i) \right), \quad i=1,\dots,\ceil*{N/2}+1, \quad j=1,\dots,N\\
\label{eq_T_is} T_\text{is}(i)&= n \quad,~s.t.~~ n:~ \frac{n(n-1)}{2} + 1 \leq i <  \frac{n(n+1)}{2} + 1,\quad n \in \mathbb{N}_+.
\end{align}
\todo[inline]{HK: Eq.\eqref{eq_basis_matrix} is somehow ugly. Maybe someone can get rid of that sum and make it nicer? I tried with 
modulo, but just could not figure it out exactly. There must be an easy way I do not see.}
Note that due to the shifting of each of the basis functions of inter-spike period $T_\text{is}$, $\bfX$ is not linear independent anymore. 
Furthermore, there will be identical basis functions and also basis functions, which do not allow for an unambiguous inter-spike period, 
if we would include all $N$ possible inter-spike periods for a signal of length $N$ instead of $\ceil*{N/2}+1$ (Fig.~\ref{fig_tau_rr_basis_functions}A). 
The reason is that in contrast to a trigonometric decomposition, where the Nyquist frequency marks a lower bound for the corresponding wave period, here 
the maximum considered inter-spike period is bounded by $T_{\text{is}}^{\text{max}} = \ceil*{N/2}+1$ (schematically illustrated in 
Fig.~\ref{fig_tau_rr_basis_functions}B).\\

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Inter spike spectra of power grid frequency data}\label{sec_power_grid}

@@

A novel type of powerspectrum, the \textit{inter-spike spectrum}, has been proposed. The method decomposes any arbitrary signal into basis functions which consist of (lagged) Dirac 
combs (DC) of different inter-spike period. The loading for each period is obtained by a regularized regression, which promotes sparsity in its solution (we chose LASSO or a sequentially 
thresholded least squares regression STLS). Since there are 
$M = \sum_{i=1}^{\ceil*{N/2}+1}i$ basis functions for a signal of length $N$ the regression can get computationally intensive for $N>1,000$. When plotting the computed loadings as a function 
of the period (or frequency) the inter spike spectrum is obtained. 
Although this novel power spectrum is superior to an ordinary FFT-based power spectrum when the signal has a spike-train-like appearance, the authors suggest that this method be considered 
as an additional source of information and not as a substitute for ordinary Fourier analysis. 
Due to the sparse regression underlying the method, there is no unique inverse of the transformation and the regularization parameter plays a crucial role and determines the 
appearance of the obtained inter spike spectrum. Moreover, similar to the Nyquist frequency barrier in the Fourier Transform which sets a lower bound for the corresponding wave period, here 
the maximum considered inter spike period is bounded by $T_{\text{is}}^{\text{max}} = \ceil*{N/2}+1$.

The invention of the proposed method has been motivated by the idea of transforming $\tau$-recurrence rate signals ($\tau$-RRs) into their frequency domain. 
This general idea \cite{Zbilut2008} allows for a frequency analysis of high dimensional systems, because the RP is a representation of the system's state space trajectory.   
The $\tau$-RR of a recurrence plot (RP) usually has a spiky shape, especially for map-like data, and the inter-spike spectrum can reliably reveal the system's dominant frequencies, 
which is not possible when Fourier transforming the $\tau$-RR or the underlying signal itself. Since the position of the peaks 
in the $\tau$-RR are not sensitive to noise, the corresponding inter spike spectrum also yields robust results in the presence of noise.

@@

@@

\clearpage

\section{Inter spike spectra for noisy R\"ossler system}

\begin{figure}[h!]
 \centering
 \includegraphics[width=0.9\textwidth]{./figures/fig_tau_rr_example_roessler_noise}
 \caption{Same as in Fig.~\ref{fig_tau_rr_example_roessler}, but here with 5\% additive Gaussian white noise on each component $x$, $y$ and $z$. The appearance of the inter spike spectra 
 in \textbf{J, K, L}, and the Fourier spectra in \textbf{M, N, P} are unaffected by the additive noise.}
\label{fig_tau_rr_example_roessler_noise}
\end{figure}

\clearpage

\section{Regularizationparameters for different sparse regression techniques}\label{sec_regularization_appendix}

\begin{figure}[h!]
 \centering
 \includegraphics[width=\textwidth]{./figures/fig_regularization}
 \caption{Pearson correlation coefficient $\rho_{\bf{s},\tilde{\bf{s}}}$ between the input signal signal $\bf{s}$ and the re-composed signal $\tilde{\bf{s}}=\bf{X}^T\hat{\boldsymbol\beta}$ as a 
 function of the regularization parameter $\alpha$ for \textbf{A} LASSO and \textbf{B} sequentially thresholded least squares (STLS) regression (see also Section \ref{sec_tau_rr_method}). 
 The input time series is of length $N=200$ and stems from the $\tau$-RR of R\"ossler system in regular dynamics.}
\label{fig_regularization}
\end{figure}
